---
title: "Divvy Bike Case Study"
author: "Andrew Sundberg"
date: "2023-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cyclistic Exercise 1st Quarter of 2020 Analysis 

###### This analysis is for case study 1 from the Google Data Analytics Certificate (Cyclistic).  It’s originally based on the case study "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman (found here: https://artscience.blog/home/divvy-dataviz-case-study). We will be using the Divvy dataset for the case study. The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: “In what ways do members and casual riders use Divvy bikes differently?”


### Install required packages


```{r}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
getwd() #displays your working directory
setwd("C:/Users/ajsun/OneDrive/Documents/Divvy_Trips_2020_Q1")
```

```{r}
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
colnames(q1_2020)

str(q1_2020)
```


```{r}
colnames(q1_2020)  #List of column names
nrow(q1_2020)  #How many rows are in data frame?
dim(q1_2020)  #Dimensions of the data frame?
head(q1_2020)  #See the first 6 rows of data frame.  Also tail(qs_raw)
str(q1_2020)  #See list of columns and data types (numeric, character, etc)
summary(q1_2020)  #Statistical summary of data. Mainly for numerics
```


```{r}
table(q1_2020$member_casual)

# Reassign to the desired values (we will go with the current 2020 labels)
q1_2020 <-  q1_2020 %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```


## Checking to make sure the proper number of observations were reassigned

```{r}
table(q1_2020$member_casual)
```


## Adding columns to list the date, day, month, and year of each ride

```{r}
q1_2020$date <- as.Date(q1_2020$started_at) #The default format is yyyy-mm-dd
q1_2020$month <- format(as.Date(q1_2020$date), "%m")
q1_2020$day <- format(as.Date(q1_2020$date), "%d")
q1_2020$year <- format(as.Date(q1_2020$date), "%Y")
q1_2020$day_of_week <- format(as.Date(q1_2020$date), "%A")
```


## Adding a "ride_length" calculation to the dataset (in seconds)

```{r}
q1_2020$ride_length <- difftime(q1_2020$ended_at,q1_2020$started_at)
```


## Convert "ride_length" from Factor to numeric so we can run calculations on the data

```{r}
is.factor(q1_2020$ride_length)
q1_2020$ride_length <- as.numeric(as.character(q1_2020$ride_length))
is.numeric(q1_2020$ride_length)
```


## The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
#### We will create a new version of the dataframe (v2) since data is being removed

```{r}
q1_2020_v2 <- q1_2020[!(q1_2020$start_station_name == "HQ QR" | q1_2020$ride_length<0),]
```


# Descriptive Analysis
## All figures in Seconds

```{r}
mean(q1_2020_v2$ride_length) #straight average (total ride length / rides)
median(q1_2020_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(q1_2020_v2$ride_length) #longest ride
min(q1_2020_v2$ride_length) #shortest ride
```


## Comparing the numbers between members and casual users

```{r}
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual, FUN = mean)
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual, FUN = median)
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual, FUN = max)
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual, FUN = min)
```


## Make sure the days of the week are in order

```{r}
q1_2020_v2$day_of_week <- ordered(q1_2020_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```


## Average ride time by each day for members vs casual users

```{r}
aggregate(q1_2020_v2$ride_length ~ q1_2020_v2$member_casual + q1_2020_v2$day_of_week, FUN = mean)
```


## Analyze ridership data by type and weekday

```{r}
q1_2020_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```


## Visualize the amount of rides by rider type

```{r}
q1_2020_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```


## Visualize for the average duration by rider type for each day of the week

```{r}
q1_2020_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

